---
#title: "Week 05. Hypothesis testing II. t-tests. Confidence intervals."
editor: source
format: 
  revealjs:
    preview-links: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
# source(here::here("materials", "00-setup.R"), local = TRUE)
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 4.5)
```

# Example 1: Effect of a drug on sleep (1) {.smaller}

::: columns
::: {.column width="50%"}
**Question**: Does the drug increase sleeping time **at the 5% level of significance**?

-   Null **hypothesis**: the drug has a zero effect
-   Alternative: the drug has a nonzero effect

**Data**:

-   sleep time for each patient before they started taking drugs (control)
-   the same after they took drugs (treatment)
-   for each patient separately, we computed a difference in sleep time between control and treatment.
-   measures are made for 10 patients
:::

::: {.column width="50%"}
::: fragment
**The estimated parameters are**:

```{r}
library(tidyverse)
library(psych)
library(knitr)
slp <- 
    sleep %>% 
    filter(group == 1) %>% 
    select(extra)
describe(slp, skew = F, ranges = F) %>% 
  kable(row.names = F)
```
:::

::: fragment
-   Can we conclude that the mean is different from zero?
:::
:::
:::

## Example 1: Effect of a drug on sleep (2)

**Can we conclude that the mean is different from zero?**

::: fragment
-   **NO!** we cannot.
:::

::: fragment
**Because:**

-   the mean difference could **result from a small sample or sample's variation.**

-   to conclude about the population, we need to take into account the sample distribution.
:::

::: fragment
We need to perform a Statistical Hypothesis Testing with:

-   The one-sample two-sided t-test.

-   Compare mean to a zero;
:::

## One-sample t-test (1): What?

One-sample t-test

::: columns
::: {.column width="65%"}
-   $H_0$ [mean is equal to the parameter `mu` specified (or zero if not specified);]{.fragment fragment-index="1"}
:::

::: {.column width="35%"}
-   $H_1$ [mean is not equal to zero;]{.fragment fragment-index="2"}
:::
:::

::: {.fragment fragment-index="4"}
-   To decide whether to reject or accept the null hypothesis, we need to rely on:
:::

::: columns
::: {.column width="60%"}
::: {.fragment fragment-index="5"}
```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,4-7,10"
t.test(slp$extra, mu = 0)
# 	One Sample t-test
# data:  slp$extra
# t = 1.3257, df = 9, p-value = 0.2176
# alternative hypothesis: true mean is not equal to 0
# 95 percent confidence interval:
#  -0.5297804  2.0297804
# sample estimates:
# mean of x 
#      0.75 
```
:::
:::

::: {.column width="40%"}
::: {.fragment fragment-index="6"}
-   The test of significance
-   The confidence interval
-   The p-value
:::
:::
:::

## One-sample t-test (2): The Test of Significance

1.  Check the $t$ statistics against the critical value $c_{df , \, 1 - \alpha / 2}$.

::: columns
::: {.column width="60%"}
```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,4"
t.test(slp$extra, mu = 0)
# 	One Sample t-test
# data:  slp$extra
# t = 1.3257, df = 9, p-value = 0.2176
# alternative hypothesis: true mean is not equal to 0
# 95 percent confidence interval:
#  -0.5297804  2.0297804
# sample estimates:
# mean of x 
#      0.75 
```
:::

::: {.column width="40%"}
::: fragment
-   $df=9$
-   $c_{0.975, 9} = `r round(qt(0.975, df = 9), 4)`$
-   if $|t| \leq c_{df , \, 1 - \alpha / 2}$ - fail to reject $H_0$;
:::
:::
:::

::: fragment
-   if $|t| > c_{df , \, 1 - \alpha / 2}$ - reject $H_0$ in favor of $H_1$ at $\alpha$ significance level;
:::

## One-sample t-test (2): Confidence intervals

2.  Check the **confidence interval** of the mean

::: columns
::: {.column width="60%"}
```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,6,7"
t.test(slp$extra, mu = 0)
# 	One Sample t-test
# data:  slp$extra
# t = 1.3257, df = 9, p-value = 0.2176
# alternative hypothesis: true mean is not equal to 0
# 95 percent confidence interval:
#  -0.5297804  2.0297804
# sample estimates:
# mean of x 
#      0.75 
```
:::

::: {.column width="40%"}
::: fragment
-   if CI includes the value from $H_0$ (0) - fail to reject $H_0$;
:::
:::
:::

::: fragment
-   if CI does not include the number from $H_0$, we reject $H_0$ in favor of $H_1$ at $\alpha$ significance level;
:::

## One-sample t-test (2): p-value

::: columns
::: {.column width="60%"}
```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,4"
t.test(slp$extra, mu = 0)
# 	One Sample t-test
# data:  slp$extra
# t = 1.3257, df = 9, p-value = 0.2176
# alternative hypothesis: true mean is not equal to 0
# 95 percent confidence interval:
#  -0.5297804  2.0297804
# sample estimates:
# mean of x 
#      0.75 
```
:::

::: {.column width="40%"}
::: fragment
-   if $p-value < \alpha$, we reject $H_0$ in favor of $H_1$ at the $p-value$ significance level;
:::
:::
:::

::: fragment
-   if $p-value \geq \alpha$, we fail to reject $H_0$;
:::

## One-sample t-test (3): Conclusion

::: columns
::: {.column width="60%"}
```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,4,7"
t.test(slp$extra, mu = 0)
# 	One Sample t-test
# data:  slp$extra
# t = 1.3257, df = 9, p-value = 0.2176
# alternative hypothesis: true mean is not equal to 0
# 95 percent confidence interval:
#  -0.5297804  2.0297804
# sample estimates:
# mean of x 
#      0.75 
```
:::

::: {.column width="40%"}
::: fragment
-   We fail to reject $H_0$
:::
:::
:::

::: fragment
-   The evidence is insufficient to conclude that the change in sleep time is significantly different from 0 at the 5% level of the sign.
:::
