---
editor: source
format: 
  revealjs:
    preview-links: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
# source(here::here("materials", "00-setup.R"), local = TRUE)
# knitr::opts_chunk$set(fig.width = 8)
# knitr::opts_chunk$set(fig.height = 4.5)
library(knitr)
library(tidyverse)
library(stringr)
library(ggplot2)
library(ggbrace)
library(ggpmisc)
library(GGally)
library(modelsummary)
library(parameters)
library(performance)
library(effectsize)
library(here)
library(patchwork)
library(equatiomatic)
ggplot2::theme_set(ggplot2::theme_minimal())
# source(here::here("materials", "00-setup.R"), local = TRUE)
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 4.5)
knitr::opts_chunk$set(fig.retina = 2)
knitr::opts_chunk$set(out.width = "100%")


get_sum_stat <- function(dta, var, group_var, dig = 2, sep = "|") {
  
  # var <- "TOTHHMEM"
  # group_var <- "Group"
  # dig <- 2
  # sep <- "|"
  
  group_var2 <- sym(group_var)
  dta %>% 
    select(var, group_var) %>% 
    pivot_longer(cols = var) %>% 
    group_by(name, !!group_var2) %>% 
    summarise(
      across(value , 
             list(n = ~ n(), 
                  mean = ~ mean(., na.rm = TRUE) %>% round(dig), 
                  sd = ~ sd(., na.rm = TRUE) %>% round(dig), 
                  se = ~ (sd(., na.rm = TRUE)/sqrt(n())) %>% round(dig)),
             .names = "{.fn}"
      )
    ) %>% 
    mutate(`Mean (SD) [SE]` = str_c(mean, " (", sd, ") [", se, "]")) %>% 
    select(-mean, -sd, -se) %>% 
    ungroup() %>% 
    pivot_wider(names_from = group_var, 
                values_from = c(n, `Mean (SD) [SE]`), 
                names_sep = sep)
}


parse_plot_ggtbl <- function(dta, parse = TRUE, drop_line = TRUE) {
  dta <- 
    dta %>%
    mutate(
      across(where(is.numeric), ~ scales::number(., accuracy = 0.01, big.mark = " "))
      )
  
  if (drop_line) {
    dta <- 
      dta %>%
      mutate(
        across(everything(), ~ifelse(row_number() == nrow(dta) - 1, "\u2026", .))
        )
  }
  
  # thm <- gridExtra::ttheme_default(padding = unit(c(1, 1), "mm"), parse = T)
  
  ggplot() +
    theme_void() +
    # annotate("table_npc",
    #          xmin = 0, xmax = 1,
    #          ymin = 0, ymax = 1,
    #          label = list(dta))+
    ggpp::geom_table(aes(
      x = 0,
      y = 0,
      label = list(dta)
    ),
    parse = parse#, 
    # table.theme = thm
    ) +
    scale_x_continuous(expand = expansion(0, 0))+ 
    scale_y_continuous(expand = expansion(0, 0))

}


kable_part <- function(x, rown_number = FALSE) {
  
  if (rown_number) {
    x <- x %>% mutate(id = row_number()) %>% select(id, everything())
  }
  xx <- x %>% slice(1:7) %>% 
    mutate(across(everything(), ~ as.character(.)))

  xx[6, seq(1, ncol(xx))] <- "$\\cdots$"
  xx %>% kable(row.names = FALSE, align = "c")
}
```

## 1. Problem setting

We would like to assess the effect of the "Conservation Reserve Program" (CPR) on the agricultural land prices.

::: incremental
Conservation Reserve Program

-   is a subsidy
-   obligates farms NOT TO GROW ANY CROPS on the enrolled land
-   pays monetary compensation in exchange;
:::

::: fragment
As this is a risk less income, it may increase the land price.

-   However, if the land is productive, such set-aside measure may reduce farmland price.
:::

```{r}
landprice_dta <- 
  alr4::MinnLand %>% 
  as_tibble()  %>%
  # filter(year == 2007) %>% 
  select(acrePrice, acres, region, 
         tillable, crpPct, productivity) %>% 
  filter(if_all(everything(), ~!is.na(.)))

fit_land_0 <- lm(
  acrePrice ~ crpPct + acres + region + tillable +  productivity, 
  landprice_dta)

fit_land_1 <- lm(
  log(acrePrice) ~ crpPct + acres + region + tillable +  productivity, 
  landprice_dta)
```

## 2. Regression model formulation {.smaller}

Hedonic price model [@palmquist1989] is an econometric approach of quantifying monetary values of differentiated characteristics of goods or services, which are subjects of economic exchange.

The theory of the hedonic price model allows us to estimate the price as a function of all characteristics of a good. Available variables in our data [@taff2007compensated] are:

::: fragment
::: columns
::: {.column width="50%"}
-   `acrePrice` - sale price in dollars per acre;
-   `crpPct` - the percentage of the farm area enrolled in CRP;
-   `acres` - size of the farm in acres;
-   `region` - region in the state Minnesota;
:::

::: {.column width="50%"}
-   `tillable` - percentage of farm that is rated arable by the assessor;
-   `productivity` - average agronomic productivity scaled 1 to 100, with larger numbers for more productive land;
:::
:::

::: fragment
EQ: **Write down a regression equation.**
:::


::: fragment
```{r}
#| results: asis
extract_eq(fit_land_0, wrap = TRUE, terms_per_line = 3 , use_coefs = FALSE)
```
:::

::: fragment
EQ: **Formulate the ex-ante hypothesis about the sign and magnitude of the coefficient XXX ?. In 1-2 sentences, explain why (based on what theory, common seance) you make this hypothesis.**
:::

:::

## 3 Exploratory data analysis {.smaller}

### 3.1 The data

```{r}
library(knitr)
set.seed(121123)
landprice_dta %>% 
  filter(crpPct  > 0) %>% 
  sample_n(15) %>%
  kable()
  # kable_part(rown_number = 15)
```

## 3.2 Summary statistics {.smaller}

```{r}
f <- All(landprice_dta) ~ N + Mean + SD + Min + Median + Max + Histogram
datasummary(f, landprice_dta)
  # get_sum_stat(
  #   c("acrePrice", "acres", "tillable", "crpPct", "productivity" ),
  #   group_var = "ggrp",
  #   sep = ""
  # ) %>% 
  # kable()

landprice_dta %>% 
  mutate(ggrp = "") %>%
  modelsummary::datasummary_skim(type = "categorical")
```

## 3.3 Regional differences {.smaller}

::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 8
#| fig-height: 6
landprice_dta %>% 
  ggplot() + 
  aes(y = acrePrice , x = region, colour = region) + 
  geom_boxplot() + 
  scale_y_log10(n.breaks = 10) +
  annotation_logticks(sides = "l") + 
  ylab("Land price, USD/acre")
```
:::

::: {.column width="25%"}
-   What kind of plat is this?
-   **What (if any) transformation of Y axis is used on the plot?**
-   What can we conclude about the land prices by region?
-   EQ: **What are the possible causes of price differences between the regions?**
-   EQ: **Does the region causes land prices to be different?**
:::
:::

## 3.4 Distributions

::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 8
#| fig-height: 6
landprice_dta %>% 
  # bind_rows(landprice_dta %>% mutate(region = "Entire state")) %>% 
  ggplot() + 
  aes(x = acrePrice , group = region, colour = region, fill = region) + 
  geom_density(alpha = 0.4) +
  # geom_boxplot() + 
  scale_x_log10(n.breaks = 10) +
  annotation_logticks(sides = "b") + 
  xlab("Land price, USD/acre")
```
:::

::: {.column width="25%"}
-   What kind of plot is this?
-   What does it show?
-   **What (if any) transformation of X axis is used on the plot?**
-   What differences and similarities does it manifest?
:::
:::

## 3.5 Scatter plots

::: columns
::: {.column width="75%"}
::: r-stack
```{r}
#| fig-width: 8
#| fig-height: 6
landprice_dta %>%
  select(-region) %>%
  ggpairs(lower = NULL, upper = NULL)
```

::: fragment
```{r}
#| fig-width: 8
#| fig-height: 6
landprice_dta %>%
  select(-region) %>%
  ggpairs(lower = NULL)
```
:::

::: fragment
```{r}
#| fig-width: 8
#| fig-height: 6
landprice_dta %>%
  select(-region) %>%
  ggpairs(lower = list(continuous = wrap(
    "smooth", alpha = 0.15, size = 0.2
  )))
```
:::
:::
:::
::: {.column width="25%"}
-   EQ: **What kind of relationship exist between `acrePrices` (land price per acre, USD) and `acres` (plot size in acres) variables?** (Linear/non linear, strong/weak correlation, positive/negative)

-   EQ: same question for any other pair of variables.
:::
:::

## 4. Estimating the regression

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1-3,7-16,18-20"
fit_land_0 <- lm(
  acrePrice ~ crpPct + acres + region + tillable +  productivity, 
  landprice_dta)
summary(fit_land_0)
# Coefficients:
#                      Estimate Std. Error t value Pr(>|t|)    
# (Intercept)           80.2822   102.3493   0.784   0.4328    
# crpPct                -6.9429     1.0727  -6.473 1.02e-10 ***
# acres                 -1.2898     0.1650  -7.818 6.00e-15 ***
# regionWest Central   995.2924    58.0391  17.149  < 2e-16 ***
# regionCentral       1950.2114    64.5434  30.215  < 2e-16 ***
# regionSouth West    1291.8278    58.3426  22.142  < 2e-16 ***
# regionSouth Central 1747.5506    62.5597  27.934  < 2e-16 ***
# regionSouth East    2040.6768    72.4923  28.150  < 2e-16 ***
# tillable               2.1123     0.9691   2.180   0.0293 *  
# productivity          23.6903     1.3300  17.812  < 2e-16 ***
# 
# Residual standard error: 1332 on 8777 degrees of freedom
# Multiple R-squared:  0.2654,	Adjusted R-squared:  0.2647 
# F-statistic: 352.4 on 9 and 8777 DF,  p-value: < 2.2e-16
```

## 6. Results interpretation {.smaller}

::: columns
::: {.column width="70%"}
```{r}
model_parameters(fit_land_0, summary = TRUE, ci = NULL) %>% 
  print_md(digits = 2)
```
:::
::: {.column width="30%"}
::: fragment
Note units of measurement: 

`acrePrice` sale price in USD per acre; 

`crpPct` is % of all farm acres enrolled under program (CRP);

`acres` size of the farm in acres;

`region` is the sub-regions in the state of Minnesota;

`tillable` is the percentage of farm that is rated arable by the assessor;

`productivity` is the average agronomic productivity scaled 1 to 100, with larger numbers for more productive land.
:::
:::
:::

::: fragment
EQ: **Write down full interpretation of and perform a hypothesis testing on each coefficient.**
:::

## 5. Checking the linearity assumption


::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 7
#| fig-height: 6
plot(x = landprice_dta$acrePrice , y = landprice_dta$acres)
```

::: fragment
Is relationship between `acrePrice` and `acres` linear?
:::

:::
::: {.column width="50%"}

::: fragment
```{r}
#| fig-width: 7
#| fig-height: 6
plot(fit_land_0, which = 1)
```
:::

::: fragment
Is the linearity assumption satisfied in our regression?
:::
:::
:::

## 4. Regression re-specification {.smaller}

```{r}
fit_land_1 <- lm(
  log(acrePrice) ~ crpPct + log(acres) + region + tillable + productivity, 
  landprice_dta)
# parameters(fit_land_1, summary = TRUE) %>% print_html(digits = 5)
```


Let us transform `acrePrice` and `acres`variables with the log transformation and estimate the regression again. The old regression is:

```{r}
#| results: asis
extract_eq(fit_land_0, wrap = TRUE, terms_per_line = 3 , use_coefs = FALSE)
```


::: fragment
**Write down the new regression equation with the transformations.**
:::


::: fragment
```{r}
#| results: asis
extract_eq(fit_land_1, wrap = TRUE, terms_per_line = 3 , use_coefs = FALSE)
```
:::

## 4.1 Results of the re-sprcified regression {.smaller}

```{r}
fit_land_1 <- lm(
  log(acrePrice) ~ crpPct + log(acres) + region + tillable + productivity, 
  landprice_dta)

# model_parameters(fit_land_1, summary = TRUE, ci = NULL) %>% 
#   print_md(digits = 2)

modelsummary(list(`log(acrePrice)` = fit_land_1), 
  output = "markdown",
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
  stars = TRUE, 
  notes = c(
    # "Note: ",
    "p-values are: '+' p < 0.1, '\\*' p < 0.05, '\\*\\*' p < 0.01, '\\*\\*\\*' p < 0.001. 
    Standard errors are reported in parentheses"))
```

## 5. Re-checking Linearity

::: columns
::: {.column width="50%"}
Before:

```{r}
#| fig-width: 7
#| fig-height: 7
plot(fit_land_0, which = 1)
```
:::

::: {.column width="50%"}
::: fragment
After the log transformation:

```{r}
#| fig-width: 7
#| fig-height: 7
plot(fit_land_1, which = 1)
```
:::
:::
:::

## 6.1 Results interpretation (with log tranformation) {.smaller}

::: columns
::: {.column width="55%"}
```{r}
# model_parameters(fit_land_1, summary = TRUE, ci = NULL) %>% 
#   print_md(digits = 3)
modelsummary(list(`log(acrePrice)` = fit_land_1), 
  output = "markdown",
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
  stars = TRUE, 
  notes = c(
    # "Note: ",
    "p-values are: '+' p < 0.1, '\\*' p < 0.05, '\\*\\*' p < 0.01, '\\*\\*\\*' p < 0.001. 
    Standard errors are reported in parentheses"))
```
:::

::: {.column width="45%"}
Log transformation requires adjusting the interpretation:

::: fragment
$\log(y) = \beta \log (x)$ is a **log-log** transformation

-   1 **% (percent)** change in $x$ causes $\beta$ **% (percent)** change in $y$ (AKA Elasticity).
:::


::: fragment
$\log(y) = \beta x$ is a **log-level** transformation

-   1 **unit** change in $x$ causes $100 \beta$ **% (percent)** change in $y$.
:::


::: fragment
$y = \beta \log(x)$ is a **level-log** transformation

-   1 **% (percent)** change in $x$ causes $\beta / 100$ **unit** in $y$.
:::

::: fragment
EQ: **Interpret selected coefficients and perform statistical hypothesis testing on them.**
:::
:::
:::

## 8. Comparing two models and concluding {.smaller}

::: columns
::: {.column width="70%"}
```{r}
modelsummary(
  list(`acrePrice` = fit_land_0, 
       `log(acrePrice)` = fit_land_1), 
  output = "kableExtra",
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
  stars = TRUE)
```

__Note:__
^^ + p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001

:::
::: {.column width="30%"}
Conclude:

::: fragment
EQ: **Which model would you choose to explain the effect of `crpPct` on the price of land?**
:::

::: fragment
EQ: **Which model best explains the relationship between X and Y?**
:::

::: fragment
**What is the effect of `crpPct` on land price?**
:::

::: fragment
**Which estimated effects are different from your ex-ante hypothesis? Provide a plausible explanation why are they different?**
:::
:::
:::
