---
editor: source
format: 
  revealjs:
    preview-links: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
# source(here::here("materials", "00-setup.R"), local = TRUE)
# knitr::opts_chunk$set(fig.width = 8)
# knitr::opts_chunk$set(fig.height = 4.5)
library(knitr)
library(tidyverse)
library(stringr)
library(ggplot2)
library(ggbrace)
library(ggpmisc)
library(GGally)
library(modelsummary)
library(parameters)
library(performance)
library(effectsize)
library(here)
library(patchwork)
library(equatiomatic)
ggplot2::theme_set(ggplot2::theme_minimal())
# source(here::here("materials", "00-setup.R"), local = TRUE)
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 4.5)
knitr::opts_chunk$set(fig.retina = 2)
knitr::opts_chunk$set(out.width = "100%")


get_sum_stat <- function(dta, var, group_var, dig = 2, sep = "|") {
  
  # var <- "TOTHHMEM"
  # group_var <- "Group"
  # dig <- 2
  # sep <- "|"
  
  group_var2 <- sym(group_var)
  dta %>% 
    select(var, group_var) %>% 
    pivot_longer(cols = var) %>% 
    group_by(name, !!group_var2) %>% 
    summarise(
      across(value , 
             list(n = ~ n(), 
                  mean = ~ mean(., na.rm = TRUE) %>% round(dig), 
                  sd = ~ sd(., na.rm = TRUE) %>% round(dig), 
                  se = ~ (sd(., na.rm = TRUE)/sqrt(n())) %>% round(dig)),
             .names = "{.fn}"
      )
    ) %>% 
    mutate(`Mean (SD) [SE]` = str_c(mean, " (", sd, ") [", se, "]")) %>% 
    select(-mean, -sd, -se) %>% 
    ungroup() %>% 
    pivot_wider(names_from = group_var, 
                values_from = c(n, `Mean (SD) [SE]`), 
                names_sep = sep)
}


parse_plot_ggtbl <- function(dta, parse = TRUE, drop_line = TRUE) {
  dta <- 
    dta %>%
    mutate(
      across(where(is.numeric), ~ scales::number(., accuracy = 0.01, big.mark = " "))
      )
  
  if (drop_line) {
    dta <- 
      dta %>%
      mutate(
        across(everything(), ~ifelse(row_number() == nrow(dta) - 1, "\u2026", .))
        )
  }
  
  # thm <- gridExtra::ttheme_default(padding = unit(c(1, 1), "mm"), parse = T)
  
  ggplot() +
    theme_void() +
    # annotate("table_npc",
    #          xmin = 0, xmax = 1,
    #          ymin = 0, ymax = 1,
    #          label = list(dta))+
    ggpp::geom_table(aes(
      x = 0,
      y = 0,
      label = list(dta)
    ),
    parse = parse#, 
    # table.theme = thm
    ) +
    scale_x_continuous(expand = expansion(0, 0))+ 
    scale_y_continuous(expand = expansion(0, 0))

}


kable_part <- function(x, rown_number = FALSE) {
  
  if (rown_number) {
    x <- x %>% mutate(id = row_number()) %>% select(id, everything())
  }
  xx <- x %>% slice(1:7) %>% 
    mutate(across(everything(), ~ as.character(.)))

  xx[6, seq(1, ncol(xx))] <- "$\\cdots$"
  xx %>% kable(row.names = FALSE, align = "c")
}
```

## 1. Problem setting {.smaller}

We would like to evaluate the causal effect of the BIMAS production intensification program on the rice output among rice-producing farmers in India.

The outcome variable is:

-   `goutput` - gross output of rice in kg per year produced on the farm.

Independent variables are:

::: incremental
-   `bimas` **variable of interest** is an intensification program. Each farmer either participates in it (yes), partially participate (mixed), or does not participate (no).

Control variables:

-   `size` the total area cultivated with rice, measured in hectares;
-   `urea` urea application in kilogram
-   `phosphate` phosphate application in kilogram
-   `hiredlabor` hired labor in hours
-   `famlabor` family labor in hours
:::

```{r}
data(RiceFarms, package = "plm")
rice_dta_0 <- 
  RiceFarms %>% 
  as_tibble() %>% 
  select(goutput, size, bimas, famlabor, hiredlabor, urea, phosphate) %>% 
  filter(if_all(everything(), ~!is.na(.))) %>% 
  filter(if_all(everything(), ~(.) !=0))

fit_rice_0 <- lm(
  goutput ~ size + bimas + famlabor + hiredlabor + urea + phosphate, 
  rice_dta_0)

fit_rice_1 <- lm(
  log(goutput) ~ log(size) + bimas + log(famlabor) + log(hiredlabor) + 
    log(urea) + log(phosphate), 
  rice_dta_0)

# modelsummary(
#   list(`acrePrice` = fit_rice_0, 
#        `log(acrePrice)` = fit_rice_1), 
#   output = "kableExtra",
#   estimate = "{estimate} ({std.error}){stars}",
#   statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
#   stars = TRUE)
```

## 2. Regression model {.smaller}

The production function is used to estimate the relationship. We consider two main production functions [see @feng2012alternative]:

::: fragment
1.  Linear [production function](https://en.wikipedia.org/wiki/Production_function), when output is a linear combination of all inputs. (**Write it down**)
:::

::: fragment
```{r}
#| results: asis
extract_eq(fit_rice_0, wrap = TRUE, terms_per_line = 3 , use_coefs = FALSE)
```
:::

::: fragment
2.  [Cobb-Douglas](https://en.wikipedia.org/wiki/Cobb%E2%80%93Douglas_production_function) production function, when the log of output is a linear combination of logged inputs. (**Write it down**)
:::

::: fragment
```{r}
#| results: asis
extract_eq(fit_rice_1, wrap = TRUE, terms_per_line = 3 , use_coefs = FALSE)
```
:::

## 2.1 Ex-ante hypothesis about the coefficients

```{r}
#| results: asis
extract_eq(fit_rice_0, wrap = TRUE, terms_per_line = 3 , use_coefs = FALSE)
```

What are the ex-ante expectations about the coefficients?

## 3.1 Exploratory data analysis (The data) {.smaller}

```{r}
library(knitr)
set.seed(121123)
rice_dta_0 %>% 
  sample_n(15) %>%
  kable()
```

## 3.2 Summary statistics {.smaller}

```{r}
f <- All(rice_dta_0) ~ N + Mean + SD + Min + Median + Max + Histogram
datasummary(f, rice_dta_0, output = "markdown")
rice_dta_0 %>% 
  modelsummary::datasummary_skim(type = "categorical", output = "markdown")
```

## 3.3 What effect BIMAS program have? {.smaller}

::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 8
#| fig-height: 6
rice_dta_0 %>% 
  ggplot() + 
  aes(y = goutput   , x = bimas , colour = bimas ) + 
  geom_boxplot() + 
  scale_y_log10(n.breaks = 10) +
  annotation_logticks(sides = "l") + 
  ylab("Rice production, kg")
```
:::

::: {.column width="25%"}
-  What kind of plot is this?
-  What can we conclude about the differences in rice output by program participation?
-  EQ: **Does the program participation cause rice output to be different?**
:::
:::

## 3.4 Distributions

::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 8
#| fig-height: 6
rice_dta_0 %>% 
  # bind_rows(landprice_dta %>% mutate(region = "Entire state")) %>% 
  ggplot() + 
  aes(x = goutput , group = bimas, colour = bimas, fill = bimas) + 
  geom_density(alpha = 0.4) +
  # geom_boxplot() + 
  scale_x_log10(n.breaks = 10) +
  annotation_logticks(sides = "b") + 
  xlab("Land price, USD/acre")
```
:::

::: {.column width="25%"}
-   What kind of plot is this?
-   What does it show?
-   What differences and similarities does it manifest?
:::
:::

## 3.5 Scatter plots

::: r-stack
```{r}
rice_dta_0 %>%
  select(-bimas) %>%
  ggpairs(lower = NULL, upper = NULL)
```

::: fragment
```{r}
rice_dta_0 %>%
  select(-bimas) %>%
  ggpairs(lower = NULL)
```
:::

::: fragment
```{r}
rice_dta_0 %>%
  select(-bimas) %>%
  ggpairs(lower = list(continuous = wrap(
    "smooth", alpha = 0.25, size = 0.75
  )))
```
:::
:::

## 4. Estimating a regression

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1-3,7-14,16-18"
fit_rice_0 <- lm(
  goutput ~ size + bimas + famlabor + hiredlabor + urea + phosphate, 
  rice_dta_0)
summary(fit_rice_0)
# Coefficients:
#               Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -107.91531   38.36178  -2.813 0.005017 ** 
# size        1824.02155  138.40695  13.179  < 2e-16 ***
# bimasyes     157.87341   83.12605   1.899 0.057866 .  
# bimasmixed  -262.07581   68.13525  -3.846 0.000129 ***
# famlabor       0.78902    0.18633   4.234 2.53e-05 ***
# hiredlabor     1.03722    0.13132   7.898 8.47e-15 ***
# urea          -0.03429    0.39242  -0.087 0.930381    
# phosphate     11.08610    0.74446  14.891  < 2e-16 ***
# 
# Residual standard error: 706.6 on 875 degrees of freedom
# Multiple R-squared:  0.8707,	Adjusted R-squared:  0.8696 
# F-statistic: 841.5 on 7 and 875 DF,  p-value: < 2.2e-16
```

## 6. Interpretation (0) {.smaller}

::: columns
::: {.column width="50%"}
```{r}
#| results: asis
extract_eq(fit_rice_0, wrap = TRUE, terms_per_line = 2 , use_coefs = TRUE)
```

::: fragment
This first regression is in `level-level` form (without any linear transformation). Thus, interpretations are straightforward:

-  1 unit change in $x$ causes $\hat \beta$ unit change in $y$.
:::
:::

::: {.column width="50%"}
```{r}
# parameters(fit_rice_0, summary = TRUE) %>% print_html( digits = 2)
# modelsummary(list(fit_rice_0), output = "kableExtra",
#              estimate = "{estimate} ({std.error}){stars}",
#              statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"))
mod_rows <- tribble(
  ~ term, ~`Linear prod. fun (level-level)`,
  "Dep. var.", "Rice output, kg"
)
attr(mod_rows, 'position') <- c(1, 2)
mod_notes <- c(
  "p-values are '+' p < 0.1, '\\*' p < 0.05, '\\*\\*' p < 0.01, '\\*\\*\\*' p < 0.001. Standard errors are reported in parentheses")
lin_prod <- 
  modelsummary(
  list(`Linear prod. fun (level-level)` = fit_rice_0#, 
       # `Cobb-Douglas prod. fun (log-log)` = fit_rice_1
       ), 
  output = "markdown",
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
  add_rows = mod_rows,
  notes = mod_notes,
  coef_map = c(
    "(Intercept)" = "(Intercept)",
    "log(size)" = "Size, HA", "size" = "Size, HA",
    "bimasyes" = "BIMAS [YES]",
    "bimasmixed" = "BIMAS [MIXED]",
    "famlabor" = "Family labor, h", "log(famlabor)" = "Family labor, h",
    "hiredlabor" = "Hired labor, h", "log(hiredlabor)" = "Hired labor, h",
    "urea" = "Urea, kg", "log(urea)" = "Urea, kg",
    "phosphate" = "Phosphate, kg", "log(phosphate)" = "Phosphate, kg"
    ),
  stars = TRUE)
lin_prod
```
:::
:::

## 6. Interpretation (1) {.smaller}

::: columns
::: {.column width="50%"}
`size` is the total area cultivated with rice, measured in hectares.

-   An increase in the farm size by 1 hectare, causes on average the gross output of rice to increase by 1824 kg, holding all other factors constant.

-   The effect of farm size is significant at the 0.1% level of significance ($H_0$ is rejected based on the p-value, which is p \< 0.001 given "\*\*\*") suggesting that a positive relationship between the farm size and output is likely to be present in the population.
:::

::: {.column width="50%"}
```{r}
lin_prod
```
:::
:::

## 6. Interpretation (2) {.smaller}

::: columns
::: {.column width="50%"}
::: {.callout-note appearance="simple"}
`bimas` is an intensification program. Each farmer either:

-   participate in it \[YES\], or partially participate \[MIXED\] or do not participate \[NO\].

-   Because the dummy variable for \[NO\] is omitted, it is implied in the intercept.

-   Variables \[YES\] and \[MIXED\] show the effect of different levels of participation **compared to NO (non-participation in BIMAS)**.
:::

-   Participating in the BIMAS on a MIXED basis reduces the gross output of rice by 262 kg compared to non-participation given that all other factors remain fixed. The effect is significant at the 5% level of significance.
:::

::: {.column width="50%"}
```{r}
lin_prod
```
:::
:::

## 6. Interpretation (3) {.smaller}

::: columns
::: {.column width="50%"}
::: {.callout-note appearance="simple"}
-   Fully participating in the BIMAS program \[YES\] yields an estimate of the coefficient of 157.8 implying on average a 158 kg increase in rice production; however, this result is insignificant at the 5% level.

-   The present functional form fails to reject $H_0$ about the significant difference of this estimate from zero concluding that such an effect may result from the random variation in the sample, rather than exist in the population.

-   To further investigate this effect, we might consider modifying the functional form or collecting more data.
:::
:::

::: {.column width="50%"}
```{r}
lin_prod
```
:::
:::

## 6. Interpretation (4) DIY {.smaller}

::: columns
::: {.column width="50%"}
::: {.callout-note appearance="simple"}
-   `urea` or `phosphate` application in kilogram;
-   `hiredlabor` or `famlabor` in hours;
:::
:::

::: {.column width="50%"}
```{r}
lin_prod
```
:::
:::

## 5. Checking the linearity assumption (1)

```{r}
plot(fit_rice_0, which = 1)
```

-   Clear signs of non-linearity.

## 4. Model re-specification {.smaller}

Estimating the Cobb-Douglas production function. Note, that $\log(\cdot)$ stands for a natural logarithm.

```{r}
#| results: asis
extract_eq(fit_rice_1, wrap = TRUE, terms_per_line = 3 , use_coefs = TRUE, coef_digits = 2)
```

## 4.1 Comparing regressions {.smaller}

::: columns
::: {.column width="80%"}
```{r}

mod_rows2 <- tribble(
  ~ term, ~`Linear (level-level)`, ~`Cobb-Douglas (log-log)`,
  "Dep. var.", "Rice output, kg", "log(Rice output, kg)"
)
attr(mod_rows2, 'position') <- c(1, 2)

both_prod <- 
  modelsummary(
  list(`Linear (level-level)` = fit_rice_0, 
       `Cobb-Douglas(log-log)` = fit_rice_1), 
  output = "markdown",
  add_rows = mod_rows2,
  notes = mod_notes,
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
  coef_map = c(
    "bimasyes" = "BIMAS [YES]",
    "bimasmixed" = "BIMAS [MIXED]",
    "size" = "Size, HA",
    "log(size)" = "log(Size, HA)", 
    "famlabor" = "Family labor, h", "log(famlabor)" = "Family labor, h",
    "hiredlabor" = "Hired labor, h", "log(hiredlabor)" = "Hired labor, h",
    "urea" = "Urea, kg", "log(urea)" = "Urea, kg",
    "phosphate" = "Phosphate, kg", "log(phosphate)" = "Phosphate, kg",
    "(Intercept)" = "(Intercept)"
    ),
  stars = TRUE)

both_prod


mod_rows3 <- tribble(
  ~ term,  ~`Cobb-Douglas (log-log)`,
  "Dep. var.", "log(Rice output, kg)"
)
attr(mod_rows3, 'position') <- c(1, 2)

cd_prod <- 
  modelsummary(
  list(#`Linear prod. fun (level-level)` = fit_rice_0, 
       `Cobb-Douglas (log-log)` = fit_rice_1), 
  output = "markdown",
  add_rows = mod_rows3,
  notes = mod_notes,
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL, gof_omit = c("AIC|BIC|Log|F|RMSE"), 
  coef_map = c(
    "bimasyes" = "BIMAS [YES]",
    "bimasmixed" = "BIMAS [MIXED]",
    "size" = "Size, HA",
    "log(size)" = "log(Size, HA)", 
    "famlabor" = "Family labor, h", "log(famlabor)" = "Family labor, h",
    "hiredlabor" = "Hired labor, h", "log(hiredlabor)" = "Hired labor, h",
    "urea" = "Urea, kg", "log(urea)" = "Urea, kg",
    "phosphate" = "Phosphate, kg", "log(phosphate)" = "Phosphate, kg",
    "(Intercept)" = "(Intercept)"
    ),
  stars = TRUE)
```
:::

::: {.column width="20%"}
-  What has changed?

-  Has any coefficient changed the sign?

-  Has any coefficient become significant / insignificant?

-  How have the goodness of fit measures (GOF) changed?
:::
:::

## 5. (Again) Checking Linearity {.smaller}

::: columns
::: {.column width="50%"}
**Linear production function**

```{r}
#| fig-width: 8
#| fig-height: 7
plot(fit_rice_0, which = 1)
```
:::

::: {.column width="50%"}
**Cobb-Douglas production function** 

::: fragment
```{r}
#| fig-width: 8
#| fig-height: 7
plot(fit_rice_1, which = 1)
```
:::

:::
:::


::: fragment
Substantial improvement in linearity.
:::

## 6. Interpretation with log (0) {.smaller}

::: columns
::: {.column width="45%"}
This regression is in the `log-log` transformation for all variables except for the dummy variables, which are in the `log-level`.

`log-log` interpretation is straightforward:

-   1 % percent increase in $x$ causes $\hat \beta$ % (percent) change in $y$.
:::

::: {.column width="55%"}
```{r}
cd_prod
```
:::
:::

## 6. Interpretation with log (1) {.smaller}

::: columns
::: {.column width="45%"}
`log-level` interpretation of the dummy variables is more complex:

-   For the purpose of the exam, please be able to explain weather the dummy variable has a positive or negative effect on $y$ and if this effect is significant.

For your information and future, please be informed that:

-   Change (increase) of a dummy variable $x$ from 0 to 1 causes $100 \cdot (e^{\hat\beta} - 1)$ % change in $y$.
:::

::: {.column width="55%"}
```{r}
cd_prod
```
:::
:::

## 6. Interpretation with log (2) {.smaller}

::: columns
::: {.column width="45%"}
`log(size)` the total area cultivated with rice, measured in hectares;

-   `log-log` transformation.

-   Increase of the farm size by 1% is associated with the 0.56% increase in the rice output, given that all other factors remain fixed.

-   These results are significant at the 5% significance level.
:::

::: {.column width="55%"}
```{r}
cd_prod
```
:::
:::

## 6. Interpretation with log (3) {.smaller}

::: columns
::: {.column width="45%"}
-   Participating in the BIMAS on a MIXED basis reduces the gross output of rice by $100 \cdot e^{-0.142}-1 = - 100 \cdot 0.132 = -13.2$ % (reduces by 13.2%) compared to non-participation given that all other factors remain fixed. The effect is significant at the 5% level of significance.

-   Participating in the BIMAS on a full basis (yes) basis increases the gross output of rice by: $100 \cdot e^{0.118}-1 = 100 \cdot 0.125 = 12.5$ % (increases by 12.2%) compared to non-participation given that all other factors remain fixed. The effect is significant at the 5% level of significance.
:::

::: {.column width="55%"}
```{r}
cd_prod
```
:::
:::

## 6. Interpretation with log (4) DIY {.smaller}

::: columns
::: {.column width="45%"}
-   `urea` or `phosphate` application in kilogram;
-   `hiredlabor` or `famlabor` in hours
:::

::: {.column width="55%"}
```{r}
cd_prod
```
:::
:::

## 8. Conclusions {.smaller}

::: columns
::: {.column width="25%"}
What is the causal effect of BIMAS program on the gross output of rice?

What is the goodness of fit of the model?

EQ: **Which model explains the relationship between inputs and outputs more accurately (with less bias)?**

EQ: **Which model has a higher explanatory power?**
:::

::: {.column width="75%"}
```{r}
both_prod
```
:::
:::
