---
editor: source
format: 
  revealjs:
    preview-links: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
source(here::here("materials", "00-setup.R"), local = TRUE)
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 4.5)


get_sum_stat <- function(dta, var, group_var, dig = 2, sep = "|") {
  
  # var <- "TOTHHMEM"
  # group_var <- "Group"
  # dig <- 2
  # sep <- "|"
  
  group_var2 <- sym(group_var)
  dta %>% 
    select(var, group_var) %>% 
    pivot_longer(cols = var) %>% 
    group_by(name, !!group_var2) %>% 
    summarise(
      across(value , 
             list(n = ~ n(), 
                  mean = ~ mean(., na.rm = TRUE) %>% round(dig), 
                  sd = ~ sd(., na.rm = TRUE) %>% round(dig), 
                  se = ~ (sd(., na.rm = TRUE)/sqrt(n())) %>% round(dig)),
             .names = "{.fn}"
      )
    ) %>% 
    mutate(`Mean (SD) [SE]` = str_c(mean, " (", sd, ") [", se, "]")) %>% 
    select(-mean, -sd, -se) %>% 
    ungroup() %>% 
    pivot_wider(names_from = group_var, 
                values_from = c(n, `Mean (SD) [SE]`), 
                names_sep = sep)
}
```

## Multiple Linear Regression

$$y = \hat {\beta}_{0} + \hat {\beta}_{1}x_1 + \hat {\beta}_{2}x_2 + \hat {\beta}_{3}x_3 + \cdots + \hat {\beta}_{k}x_k + \hat u$$

::: incremental
-   $\hat u$ - Error term, or disturbance containing factors other than $x_1, x_2, \cdots, x_k$ that affect $y$

-   $\hat {\beta}_{0}$ - intercept (constant term);

-   $\hat {\beta}_{1}$, $\hat {\beta}_{2}$, $\hat {\beta}_{k}$ - coefficients / parameters associated with $x_1$, $x_2$, ... $x_k$;

-   $k$ - entire set of independent variables;
:::

## Motivation befind MLR

::: incremental
-   To incorporate more explanatory factors into a model;

-   Explicitly hold fixed (some) other factors;

-   Allow for more flexible functional forms;

-   To correct the selection bias problem in the absence of (poor quality of) the RCT;

-   To establish the ceteris paribus in observational data;
:::

## Interpretation

-   The multiple linear regression shows the effect of each variable, holding other explanatory variables fixed;

::: fragment
```{r}
#| echo: false
#| fig-asp: 0.15
here("materials", "img", "ex2-interpretation.png") %>%
  image_read() %>%
  image_ggplot(interpolate = TRUE)
```
:::

::: fragment
::: callout-important
### We assume that all unobserved factors do not change if the explanatory variables are changed.
:::
:::

## Example 1. Wage equation

```{r}
#| echo: false
#| fig-asp: 0.25
#| 
here("materials", "img", "ex2-exmple-00.png") %>%
  image_read() %>%
  image_ggplot(interpolate = TRUE)
```

## Example 2. Average test scores and spending

```{r}
#| echo: false
#| fig-asp: 0.2
here("materials", "img", "ex2-exmple-01.png") %>%
  image_read() %>%
  image_ggplot(interpolate = TRUE)
```

## Example 3. CEO salary, sales and CEO tenure

```{r}
#| echo: false
#| fig-asp: 0.15
here("materials", "img", "ex2-exmple-03.png") %>%
  image_read() %>%
  image_ggplot(interpolate = TRUE)
```

::: incremental
-   Model assumes a constant elasticity relationship between CEO salary and the sales of his or her firm.

-   Model assumes a quadratic relationship between CEO salary and his or her tenure with the firm.
:::

## Mechanics

$$y = \hat {\beta}_{0} + \hat {\beta}_{1}x_1 + \hat {\beta}_{2}x_2 + \hat {\beta}_{3}x_3 + \cdots + \hat {\beta}_{k}x_k + \hat u$$

::: incremental
1.  MLR uses the Ordinary Least Squares (OLS) method to **estimate regression parameters** $\beta$.

    -   Such parameters $\beta$ are found that the sum of squared errors is minimal.

2.  As any estimate of $\beta$ is based on a sample, we need to relay on the **inference** to test a hypothesis about the true coefficients in the population.
:::

# Example MLR (1). Effect of fuel taxes on petrol consumption

We use data on motor fuel consumption and related variables, for the year 2001. Each observation represents a state in the United States. (based on the data from @weisberg2005a). Variables are:

-   $Tax$ : Gasoline state tax rate, cents per liter
-   $Dlic$ : The number of licensed drivers per 1000 population over the age of 16;
-   $Income$ : in 1000 USD Per capita personal income (year 2000)
-   $Miles$ : Miles of Federal-aid highway miles in the state
-   $Fuel$ : Gasoline consumption per capita (liter)

## 1. Problem formulation {.smaller}

How do fuel taxes affect gasoline consumption?:

::: fragment
### 2. Regression model and hypothesis

Write a regression equation:

::: fragment
$$Fuel = \hat\beta_0 + \hat\beta_1  Tax + \hat\beta_2  Dlic + \hat\beta_3 Income + \hat\beta_4  Miles + \hat{u}$$
:::

Discuss, what effects can we expect from change in each variable $x$ on $y$?

::: fragment
Hypotheses:

-   $\hat\beta_0$ [$>0$ because on average some consumption should occur]{.fragment}
-   $\hat\beta_1$ [$<0$ taxes make flue more expensive leading people to consume less]{.fragment}
-   $\hat\beta_2$ [$>0$ more people with driving licences increase consumption of flue]{.fragment}
-   $\hat\beta_3$ [$>0$ higher income relaxes budget constraints allowing to drive more]{.fragment}
-   $\hat\beta_4$ [$>0$ a larger road network makes people drive more]{.fragment}
:::
:::

```{r}
fule_cons <- as_tibble(alr4::fuel2001)
fule_cons <- mutate(fule_cons, 
                    Dlic = Drivers / (Pop/ 1000),
                    Fuel = FuelC / Pop * 1000,
                    Income = Income / 1000,
                    logMiles = log(Miles))
fule_cons <- select(fule_cons, Tax, Dlic, Income, logMiles, Fuel)
fit_fule_1 <- lm(Fuel ~  Tax + Dlic + Income + logMiles, fule_cons)
```

## Exploratory data analysis (1) {.smaller}

::: columns
::: {.column width="50%"}
```{r}
psych::describe(fule_cons, fast = T) %>% 
  rownames_to_column() %>% 
  tidyr::pivot_longer(-1) %>%
  tidyr::pivot_wider(names_from = name, values_from = value) %>% 
  select(-vars, -n, -min, -max, -range) %>% 
  kable(digits = 2)
```

::: fragment
**Variable of interest:** `Tax`.

**Outcome variable:** `Fule`.

::: {.callout-note appearance="simple"}
### What kind of relationship exists between Tax and Fuel?
:::
:::
:::

::: {.column width="50%"}
```{r}
#| fig-width: 10
#| fig-height: 9
ggpairs(fule_cons)
```
:::
:::

## Exploratory data analysis (2)

### What kind of relationship exists between Tax and Fuel?

::: r-stack
```{r}
#| fig-width: 8
#| fig-height: 4
fule_cons %>% 
  ggplot() +
  aes(y = Fuel, x = Tax) + 
  geom_point() + 
  xlab("Tax, cents per liter") + 
  ylab("Fuel consumption, liter per capita")
```

::: fragment
```{r}
#| fig-width: 8
#| fig-height: 4
fule_cons %>% 
  ggplot() +
  aes(y = Fuel, x = Tax) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  xlab("Tax, cents per liter") + 
  ylab("Fuel consumption, liter per capita")
```
:::
:::

## 4. Estimating a Simple Linear Reg.

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,5-8"
fit_fule_0 <- lm(Fuel ~  Tax, fule_cons)
summary(fit_fule_0)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  715.485     55.770  12.829   <2e-16 ***
# Tax           -5.078      2.701  -1.881    0.066 .  
# Residual standard error: 86.79 on 49 degrees of freedom
# Multiple R-squared:  0.06731,	Adjusted R-squared:  0.04828 
# F-statistic: 3.536 on 1 and 49 DF,  p-value: 0.06599
```

### What does it show?

::: incremental
-   The estaimted effect of taxes is -5.1. Full answer:
    -   A unit change in tax level (1 cent increase in taxes per liter)
    -   **reduces** fuel consumption by 5.1 liters per person
    -   assuming all other factors are fixed.
-   However, p-value \> 0.05. The same effect of taxes may not hold for the population.
-   Is this the true **causal effect**? Vote .... Yes/no?
-   NO! There is no Ceteris Paribus here. We do not control for other things with a **Simple** regression. We need a **Multiple regression**
:::

## 4. Estimating the Multiple Linear Regression

Multiple linear regression helps to reveal the true causal effect of `Tax` by

-   controlling all other things.
-   making them equal when the tax changes by one unit.

::: fragment
```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,5-6,8-9"
fit_fule_1 <- lm(Fuel ~  Tax + Dlic + Income + logMiles, fule_cons)
summary(fit_fule_1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 154.1928   194.9062   0.791 0.432938    
# Tax          -4.2280     2.0301  -2.083 0.042873 *  
# Dlic          0.4719     0.1285   3.672 0.000626 ***
# Income       -6.1353     2.1936  -2.797 0.007508 ** 
# logMiles     26.7552     9.3374   2.865 0.006259 ** 
# 
# Residual standard error: 64.89 on 46 degrees of freedom
# Multiple R-squared:  0.5105,	Adjusted R-squared:  0.4679 
# F-statistic: 11.99 on 4 and 46 DF,  p-value: 9.331e-07
```
:::

## 6. Inspecting and interpreting the results

Here we need to:

-   perform the HT about estimated coefficients;

-   interpretation of their economic meaning;

-   compare estimates with ex-ante expectations.

## The Intercept {.smaller}

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,5"
fit_fule_1 <- lm(Fuel ~  Tax + Dlic + Income + logMiles, fule_cons)
summary(fit_fule_1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 154.1928   194.9062   0.791 0.432938    
# Tax          -4.2280     2.0301  -2.083 0.042873 *  
# Dlic          0.4719     0.1285   3.672 0.000626 ***
# Income       -6.1353     2.1936  -2.797 0.007508 ** 
# logMiles     26.7552     9.3374   2.865 0.006259 ** 
# 
# Residual standard error: 64.89 on 46 degrees of freedom
# Multiple R-squared:  0.5105,	Adjusted R-squared:  0.4679 
# F-statistic: 11.99 on 4 and 46 DF,  p-value: 9.331e-07
```

::: incremental
-   Positive but insignificant at the 5% level. (p-value = 0.432).

-   Contradicts our ex-ante hypothesis about some basic level of fuel consumption;

-   Suggests that the absence of a road network could drive fuel consumption to zero.
:::

## The Slope of Taxes {.smaller}

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,6"
fit_fule_1 <- lm(Fuel ~  Tax + Dlic + Income + logMiles, fule_cons)
summary(fit_fule_1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 154.1928   194.9062   0.791 0.432938    
# Tax          -4.2280     2.0301  -2.083 0.042873 *  
# Dlic          0.4719     0.1285   3.672 0.000626 ***
# Income       -6.1353     2.1936  -2.797 0.007508 ** 
# logMiles     26.7552     9.3374   2.865 0.006259 ** 
# 
# Residual standard error: 64.89 on 46 degrees of freedom
# Multiple R-squared:  0.5105,	Adjusted R-squared:  0.4679 
# F-statistic: 11.99 on 4 and 46 DF,  p-value: 9.331e-07
```

::: incremental
-   Negative and significant at the 5% level (p-value = 0.043).

-   One cent increase in taxes per liter causes flue consumption to reduce by -4.2 liter per capita, given that all other factors remain fixed.

    -   Note: $Tax$ Gasoline state tax rate, cents per liter and $Fuel$ Gasoline consumption per capita (liter)

-   This is in line with our ex-ante hypothesis.
:::

## The Slope of Income {.smaller}

Note: $Income$ in 1000 USD Per capita personal income (the year 2000)

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,8"
fit_fule_1 <- lm(Fuel ~  Tax + Dlic + Income + logMiles, fule_cons)
summary(fit_fule_1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 154.1928   194.9062   0.791 0.432938    
# Tax          -4.2280     2.0301  -2.083 0.042873 *  
# Dlic          0.4719     0.1285   3.672 0.000626 ***
# Income       -6.1353     2.1936  -2.797 0.007508 ** 
# logMiles     26.7552     9.3374   2.865 0.006259 ** 
# 
# Residual standard error: 64.89 on 46 degrees of freedom
# Multiple R-squared:  0.5105,	Adjusted R-squared:  0.4679 
# F-statistic: 11.99 on 4 and 46 DF,  p-value: 9.331e-07
```

::: incremental
-   Negative and significant at the 5% level (p-value = 0.043).

-   An increase in income by 1000 USD per person causes fuel consumption to reduce by -6.1 liter per capita, ceteris paribus.

-   This contradicts our ex-ante hypothesis. Why?
:::

## The Slope of logMiles {.smaller}

Note: $logMiles$ is the logarithm of miles. Interpretation differs because this is a level-log transformation

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "1,2,9"
fit_fule_1 <- lm(Fuel ~  Tax + Dlic + Income + logMiles, fule_cons)
summary(fit_fule_1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 154.1928   194.9062   0.791 0.432938    
# Tax          -4.2280     2.0301  -2.083 0.042873 *  
# Dlic          0.4719     0.1285   3.672 0.000626 ***
# Income       -6.1353     2.1936  -2.797 0.007508 ** 
# logMiles     26.7552     9.3374   2.865 0.006259 ** 
# 
# Residual standard error: 64.89 on 46 degrees of freedom
# Multiple R-squared:  0.5105,	Adjusted R-squared:  0.4679 
# F-statistic: 11.99 on 4 and 46 DF,  p-value: 9.331e-07
```

::: incremental
-   Positive and significant at the 5% level (p-value = 0.043).

-   An increase in road network by 1% causes fuel consumption to increase by 26.7/100 (0.26) liter per capita, given ceteris paribus.
:::
