---
#title: "t-test"
editor: source
format: 
  revealjs:
    preview-links: true
    incremental: false
editor_options: 
  chunk_output_type: consoles
---

# t-test

Most commonly used statistical test for hypothesis testing about the estimated parameter (e.g., $\widehat{\text{mean}}$, but also other estimates).

**The purpose:**

Infer:

::: incremental
-   whether or not the population parameter $\mu$ (mean, or else)

-   is different from an arbitrary number (usually zero: $\mu_{0} = 0$)

-   based on our sample estimates of

    -   the parameter ($\widehat{\text{mean}}$),
    -   the variance ($\text{se}$ or $s$), and
    -   sample size $n$.
:::

## t-tests

In principal, there are:

::: incremental
-   one-sample t-test: compares mean to a number

-   two-sample t-test: compares means of two independent samples

    -   for equal and unequal variances between samples
:::

## one-sample t-test: comparing mean to a number

### The algorithm

1.  **Step 1.** Hypothesis formulation;

2.  **Step 2.** Calculating test statistic;

3.  **Step 3.** Setting a decision rule for rejecting $H_0$;

4.  **Step 4.** Making a conclusion;

## Step 1. Hypothesis formulation

::: columns
::: {.column width="50%"}
::: fragment
### $H_0$ Null Hypothesis

The population mean equals to zero $\mu_{0}$

-   $H_0 : \mu = \mu_{0}$
:::
:::

::: {.column width="50%"}
::: fragment
### $H_1$ Alternative Hypothesis

The population mean is not zero $\mu_{0}$

::: incremental
-   $H_1: \mu \ne \mu_{0}$

-   $H_1: \mu < \mu_{0}$

-   $H_1: \mu > \mu_{0}$
:::
:::
:::
:::

## Step 2. Calculating test statistic

::: columns
::: {.column width="60%"}
::: fragment
### t-statistics:

$$t = (\hat\mu - \mu_0) / \text{se} =  \frac{\sqrt{n} ( \hat\mu - \mu_0)}{s}\,,$$
:::

::: fragment
where:

-   $t$ - t-statistics;

-   $\hat\mu$ - estimated mean and $\mu_0$ is an arbitrary number;

-   $n$ - number of observation and $s$ - sample standard deviation;
:::
:::

::: {.column width="40%"}
::: fragment
Sometimes, t-statistics is defined as:

$$t = (\hat\mu -  \mu_0) / se(\hat\mu) \, ,$$

where:

-   $se(\hat\mu)$ is the standard error of the estimated coefficient $\hat\mu$

-   $se(\hat\mu) = s / \sqrt{n}$;
:::
:::
:::

## Step 2. Example

::: columns
::: {.column width="60%"}
::: incremental
Sample size: $n = 25$

Estimates mean: $\hat \mu = -2.2$

Sample SD: $s = 3.3$

-   Compute the t-statistics for the mean difference from zero.

-   $\mu_0 = 0$?

-   $t = \frac{\sqrt{25} ( -2.2 - 0)}{3.3} \\ = 5 \cdot (-1/3) \\= -1.667$
:::
:::

::: {.column width="40%"}
::: fragment
::: incremental
-   Compute the t-statistics for the mean difference from -5.5.

-   $\mu_0 = -5.5$?

-   $t = \frac{\sqrt{25} ( -2.2 + 5.5)}{3.3} \\ = 5 \cdot 1 \\ = 5$
:::
:::
:::
:::

## Step 3. Setting a decision rule (1)

To reject $H_0$, we compare $t$ to a **critical value** $c_{df, \text{CL}}$, where:

-   $\text{CL}$ is the confidence level (opposite to the significance level: $1-\alpha$)

-   $df$ - degrees of freedom

## Step 3. Setting a decision rule (2)

**Critical value**, is the **minimum values of the t-statistics** at which we **can reject** $H_0$ at the given **level of significance**.

-   Critical value depends on the alternative hypothesis:

-   when $H_1: \mu \ne \mu_{0}$ (two-sided t-test),

    -   $\text{CL} = 1 - \alpha / 2$ then $c_{df,\, 1 - \alpha / 2}$;

-   $H_1: \mu < \mu_{0}$ or $H_1: \mu > \mu_{0}$ (one-sided t-test),

    -   $\text{CL} = 1 - \alpha$ then $c_{df,\, 1 - \alpha}$; :::

## Step 3. Setting a decision rule (3)

::: incremental
Critical value also depends on:

-   the number of degrees of freedom $df = n - k$:

    -   $n$ is the number of observation in the sample and

    -   $k$ is the number of parameters

        -   $k=1$ for one-sample t-test

        -   $k=2$ for two-sample t-test

-   level of significance $\alpha$;

    -   Conventionally $\alpha = 0.05$
:::

## Step 3. Setting a decision rule (4)

::: incremental
We reject $H_0: \mu = \mu_0$ in favor of $H_1: \mu \neq \mu_{0}$ **only if**:

-   $|t| > c_{df, 1 - \alpha / 2}$ for two-sided t-test $H_1: \mu \ne \mu_{0}$

-   $|t| > c_{df, 1 - \alpha}$ for one-sided t-test $H_1: \mu > \mu_{0}$ or $H_1: \mu > \mu_{0}$
:::

## Step 3. Setting a decision rule (4) {.smaller}

::: columns
::: {.column width="30%"}
Finding the critical value is possible from the quantiles of the t-distribution.

-   $df = n - k$;
-   $\text{CL} = 1 - \alpha / 2$ for $H_1: \mu \ne \mu_{0}$
-   $\text{CL} = 1 - \alpha$ for $H_1: \mu > \mu_{0}$ or $H_1: \mu > \mu_{0}$
:::

::: {.column width="70%"}
```{r eval = TRUE, echo = FALSE, results = "asis"}
library(purrr)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)
critical_tbl <- 
  c(1, 5, 10, 20, 23, 24, 25, 26, 27, 28, 29, 30, 50, 150) %>% 
  map_dfr(~{
    c(0.95, 0.975, 0.99, 0.995) %>% 
      set_names(qt(., .x) %>% round(4), .) %>% 
      enframe() %>% 
      mutate(df = .x)
  }) %>% 
  pivot_wider(names_from = "name", values_from = "value")  %>% 
  knitr::kable()

c(1, 5, 10, 25, 40, 50, 75, 100, 150, 300, 1000, 10000, 1000000) %>% 
  map_dfr(~{
    c(0.9, 0.95, 0.975, 0.99, 0.995) %>% 
      set_names(qt(., .x) %>% round(4), .) %>% 
      enframe() %>% 
      mutate(df = .x)
  }) %>% 
  pivot_wider(names_from = "name", values_from = "value")  %>% 
  knitr::kable()
```
:::
:::

## Step 3. Example (1) {.smaller}

::: columns
::: {.column width="50%"}
Sample size: $n = 25$; $df = 25 - 1 = 24$

::: fragment
Is the population mean different from zero at a 1% level of significance?

::: incremental
-   $H_1: \mu \ne 0$
-   $1 - \alpha / 2 = 1 - 0.01 / 2 = 0.995$
-   $c_{24, \, 1 - 0.01/ 2} = 2.7969$
:::
:::
:::

::: {.column .smaller width="50%"}
```{r eval = TRUE, echo = FALSE, results = "asis"}
critical_tbl
```
:::
:::

## Step 3. Example (2) {.smaller}

::: columns
::: {.column width="50%"}
Sample size: $n = 25$; $df = 25 - 1 = 24$

::: fragment
Is the population mean different from zero at a 5% level of significance?

::: incremental
-   $1 - \alpha / 2 = 1 - 0.05 / 2 = 0.975$

-   $c_{24, \, 1 - 0.05/ 2} = 2.0639$
:::
:::
:::

::: {.column width="50%"}
```{r eval = TRUE, echo = FALSE, results = "asis"}
critical_tbl
```
:::
:::

## Step 3. Example (3) {.smaller}

::: columns
::: {.column width="50%"}
Sample size: $n = 25$; $df = 25 - 1 = 24$.

::: fragment
Is the population mean different from zero at 10% level of significance?

::: incremental
-   $1 - \alpha / 2 = 1 - 0.1 / 2 = 0.95$

-   $c_{24, \, 1 - 0.1/ 2} = 1.7109$
:::
:::
:::

::: {.column width="50%"}
```{r eval = TRUE, echo = FALSE, results = "asis"}
critical_tbl
```
:::
:::

## Step 4. Make a conclusion

Concluding either or:

::: incremental
-   We reject $H_0$

    -   at least at the $\alpha$ level of significance.
    -   The population mean is significantly different from $\mu_0$.

-   We fail to reject (we accept) $H_0$.

    -   Available data do not provide sufficient evidence to reject $H_0$ at the $\alpha$ level of significance.
    -   The population mean is statistically indifferent from $\mu_0$.
:::

## Step 4. Example (1)

Sample size: $n = 25$. Estimates mean: $\hat \mu = -2.2$. Sample SD: $s = 3.3$

::: fragment
Is the population mean different from zero ($\mu_0 = 0$) at 5% level of significance?

::: incremental
-   $t = \frac{\sqrt{25} ( -2.2 - 0)}{3.3} = -1.667$

-   $c_{24, \, 1 - 0.05/ 2} = 2.0639$

-   $|t| < c_{24, \, 1 - 0.01/ 2}$

-   $|-1.667| < 2.0639$

-   Conclusion:

    -   Fail to reject (we accept) $H_0$.
    -   the population mean is statistically **indifferent** from $\mu_0 = 0$.
:::
:::

## Step 4. Example (2)

::: fragment
Is the population mean different from -5.5 ($\mu_0 = -5.5$) at the 1% level of significance?

::: incremental
-   $\mu_0 = -5.5$?

-   $t = \frac{\sqrt{25} ( -2.2 + 5.5)}{3.3} = 5$

-   $c_{24, \, 1 - 0.05/ 2} = 2.7969$

-   $|t| > c_{24, \, 1 - 0.01/ 2}$

-   $|5| > 2.7969$

-   Conclusion:

    -   Reject $H_0$.
    -   the population mean is statistically different from $\mu_0 = -5.5$ at least at the 1% level of significance.
:::
:::

## Homework (exam) examples (1) {.smaller}

::: columns
::: {.column width="50%"}
Calculate the t-statistics and find corresponding critical values.

-   Conclude weather the H0 could be rejected in the following hypothesis testing problems.

-   Compare your calculations to the solutions.

```{r}
tibble(
  `$\\hat \\mu$` = c(4, 2, 3.5, 8, 1.2),
  `s` = c(1, 1.2, 3.1, 4, 3.4),
  `$\\mu_0$` = c(3, 0.8, 2.8, 16, .2),
  n = c(10, 8, 150, 21, 53),
  `$\\alpha$` = c(0.01, 0.05, 0.01, 0.1, 0.05)
) %>%
  mutate(
    `$t$` = round(sqrt(n) * (`$\\hat \\mu$` - `$\\mu_0$`) / `s`, 3),
    `$c_{df, \\text{CL}}$` = round(qt(1 - `$\\alpha$` / 2, n), 3)
  ) %>% 
  kable()
```
:::

::: {.column width="50%"}
```{r table, eval = TRUE, echo = FALSE, results = "asis"}
c(1, 5, 6, 7, 8, 9, 10, 20, 25, 30, 50, 80, 100, 150) %>% 
  map_dfr(~{
    c(0.95, 0.975, 0.99, 0.995) %>% 
      set_names(qt(., .x) %>% round(4), .) %>% 
      enframe() %>% 
      mutate(df = .x)
  }) %>% 
  pivot_wider(names_from = "name", values_from = "value")  %>% 
  knitr::kable(caption = "Critical values of the t-distribution.")
```
:::
:::
